A Practical Optional Type System for Clojure

### 2012-10-16 dissertation/experiments.tex
Lots of stuff

Monads[] are an interesting control structure probably most recognisable from its inclusion
in the statically-typed language Haskell[].

I chose to port a subset of the Clojure library \emph{algo.monads} to Typed Clojure. The library
provides support for several kinds of monads.

@@

A monad is defined using the macro \lstinline|defmonad|. The macro expands to code that binds a var to a hash-map
with the aforementioned keys. For example, listing \ref{lst:identitymdef} expands to 
the code in listing \ref{lst:identitymexpand}.

\begin{lstlisting}[caption=Type for identity monad, label=lst:identitymtype]
(ann identity-m
     '{:m-bind (All [x y]
                 [x [x -> y] -> y])
       :m-result (All [x]
                   [x -> x])
       :m-zero '::undefined
       :m-plus '::undefined})
\end{lstlisting}

Typed Clojure operates directly on fully macro-expanded forms, so it is unsurprising that the type
annotation for \lstinline|identity-m| reflects its macro-expansion (listing \ref{lst:identitymtype}).
The \lstinline|:m-bind| entry is comparable to Haskell's monadic \emph{bind},
and the \lstinline|:m-result| entry is comparable to Haskell's monadic \emph{return}.

\begin{lstlisting}[caption=Maybe monad definition, label=lst:maybemdef]
(defmonad maybe-m
  "Monad describing computations with possible failures. Failure is
  represented by nil, any other value is considered valid. As soon as
  a step returns nil, the whole computation will yield nil as well."
  [m-zero   nil
   m-result (fn m-result-maybe [v] 
              v)
   m-bind   (fn m-bind-maybe [mv f]
              (if (nil? mv) nil (f mv)))
   m-plus   (fn m-plus-maybe [& mvs]
              (first 
                (filter #(not (nil? a))) mvs))
   ])
\end{lstlisting}

Instead of explaining the identity monad further, we show a more interesting monad, the maybe monad.
Its untyped definition is given in listing \ref{lst:maybemdef}.

\begin{lstlisting}[caption=Typed maybe monad definition, label=lst:maybemtyped]
(ann maybe-m
    '{:m-bind (All [x y]
                [(U nil x) [x -> (U nil y)] -> (U nil y)])
      :m-result (All [x]
                  [x -> (U nil x)])
      :m-zero nil
      :m-plus (All [x]
                [(U nil x) * -> (U nil x)])})
(defmonad maybe-m
  "Monad describing computations with possible failures. Failure is
  represented by nil, any other value is considered valid. As soon as
  a step returns nil, the whole computation will yield nil as well."
  [m-zero   nil
   m-result (ann-form 
              (fn m-result-maybe [v] v)
              (All [x] 
                [x -> (U nil x)]))
   m-bind   (ann-form 
              (fn m-bind-maybe [mv f]
                (if (nil? mv) nil (f mv)))
              (All [x y]
                [(U nil x) [x -> x] -> (U nil x)]))
   m-plus   (ann-form 
              (fn m-plus-maybe [& mvs]
                (first 
                  (filter
                    (ann-form 
                      #(not (nil? %))
                      [(U nil x) -> boolean :filters {:then (is x 0)}])
                    mvs)))
              (All [x]
                [(U nil x) * -> (U nil x)]))
   ])
\end{lstlisting}

A significant difference between typing Haskell style monads in Haskell and typing monads in \emph{algo.monads} with Typed Clojure
is that monads in Haskell are represented by type classes, while Typed Clojure does not support higher-kinded
constructs.

>>>>

Monads are an interesting control structure most used in functional programming languages,
as described by Wadler\cite{Wad95}.
They are most recognisable from its inclusion in the statically-typed language Haskell\cite{Mar10},
where monads are relied on for many features including global state, file output, and exceptions.

I chose to port the Clojure Contrib library \emph{algo.monads}\cite{Hic12} to Typed Clojure. The library
provides several kinds of monads, monad transformers, and monadic functions.
Macros are used to provide pleasant syntax for consumers of the library.

\subsection{Monad Definitions}
\label{sec:monaddef}

@@

A monad is defined using the macro \lstinline|defmonad|. The macro expands to code that binds a var to a hash-map
with the aforementioned keys.

\begin{lstlisting}[caption=Type for identity monad, label=lst:identitymtype]
(def-alias Undefined '::undefined)
(ann identity-m
     '{:m-bind (All [x y]
                 [x [x -> y] -> y])
       :m-result (All [x]
                   [x -> x])
       :m-zero Undefined
       :m-plus Undefined})
\end{lstlisting}

A direct typing of the identity monad would look like listing \ref{lst:identitymtype}.
This signature assigns the expected type for the var \lstinline|identity-m| 
to be a heterogeneous map type with the minimum entries for monads: \lstinline|:m-bind|
and \lstinline|:m-result|. The syntax for heterogeneous map types requires a \lstinline|'| prefix,
and any number of key-value pairs are given between curly braces. Heterogeneous map keys must be keywords,
so the syntax is made more convenient by omitting the usually required \lstinline|'| prefix
for keyword types.

Monad types in languages with advanced type systems are often abstracted using type constructors.
This allows reasoning about monadic code while keeping the particular monad abstract, which is a desirable result,
so Typed Clojure was extended to support user definable type constructors.

\begin{lstlisting}[caption=An abstract definition of a monad., label=lst:monadalias]
(def-alias Monad 
  (TFn [[m :kind (TFn [[x :variance :covariant]] Any)]]
    '{:m-bind (All [x y]
                [(m x) [x -> (m y)] -> (m y)])
      :m-result (All [x]
                  [x -> (m x)])
      :m-zero Undefined
      :m-plus Undefined}))
\end{lstlisting}

Listing \ref{lst:monadalias} captures the abstract definition of a monad.
A \lstinline|Monad| is a type constructor parameterised by \lstinline|m|, which is
a type constructor taking a single argument \lstinline|x| (a type) and returning a type 
(written \lstinline|Any|)\footnote{Haskell-like syntax is helpful here, the kind \lstinline|(TFn [[x :variance :covariant]] Any)|
is approximately \lstinline|* -> *|. A more streamlined syntax is planned for future work.}.
The body of the type constructor uses \lstinline|m| abstractly. Typed Clojure
ensures the correct number of arguments are passed and recognises the declared variances
for each parameter, In this case the first argument of \lstinline|m| is declared
a covariant position. When instantiated, \lstinline|m| must also be a type operator
of one covariant argument returning a type.

\begin{lstlisting}[caption=Identity monad using user defined type constructors, label=lst:identityctors]
(ann identity-m (Monad (TFn [[x :variance :covariant]] x)))
\end{lstlisting}

We can now express the type of the identity monad more abstractly (listing \ref{lst:identityctors}).
The purpose of the monad seems more apparent just from reading its type.
In this case, the fact that identity monad has no effect is reflected
by its type constructor returning exactly its argument.

\begin{lstlisting}[caption=Type checked identity monad definition, label=lst:identitydeftyped]
(defmonad identity-m
   "Monad describing plain computations. This monad does in fact nothing
    at all. It is useful for testing, for combination with monad
    transformers, and for code that is parameterized with a monad."
  [m-result identity
   m-bind   (ann-form
              (fn m-result-id [mv f]
                (f mv))
              (All [x y]
                [x [x -> y] -> y]))
  ])
\end{lstlisting}

The final type checked definition for the identity monad is given in listing \ref{lst:identitydeftyped}.
In this case, just the monadic bind required annotation.

\begin{lstlisting}[caption=Several monad types, label=lst:severalmonads]
; Maybe monad
(ann maybe-m (MonadPlusZero
               (TFn [[x :variance :covariant]] 
                 (U nil x))))

; Sequence monad (called "list monad" in Haskell)
(ann sequence-m (MonadPlusZero 
                  (TFn [[x :variance :covariant]]
                    (Seqable x))))

; State monad
(def-alias State
  (TFn [[r :variance :covariant]
        [s :variance :invariant]]
    [s -> '[r s]]))
(ann state-m (All [s]
               (Monad (TFn [[x :variance :covariant]]
                        (State x s)))))
\end{lstlisting}

Listing \ref{lst:severalmonads} shows several monad types. The alias \lstinline|MonadPlusZero|
is identical to \lstinline|Monad|, except both \lstinline|:m-zero| and \lstinline|:m-plus| are defined.
The definitions were type checked by adding type annotations in appropriate places, in a similar fashion
to the identity monad. A significant obstacle was discovered in while type checking monad definitions,
relating to filtering sequences with negative type information, discussed in section \ref{sec:filterneg}.

\subsection{Monad Transformer Definitions}

The initial motivation for adding user defined type constructors to Typed Clojure
(as discussed in section \ref{sec:monaddef}) was to type check
monad transformer definitions.
Monad transformers are always parameterised by a monad type constructor, and should work for all
monad type constructors. Keeping the monad type constructor abstract allows us to determine
whether a monad transformer works \emph{for any} monads.

### 2012-10-16 dissertation/intro.tex
Add implementation
Lots of stuff

The Clojure programming language is a dynamically typed language running on the Java Virtual Machine. 
It emphasises functional programming with immutable data structures
and provides direct interoperability with languages on the Java Virtual Machine.
Clojure also is also a family of languages with similar semantics, referred to as Clojure dialects.
There are several Clojure dialects, notably the aforementioned Clojure for the Java Virtual Machine,
ClojureCLR which targets the Common Language Runtime, and ClojureScript which compiles to Javascript.
We refer to Clojure the language as \emph{Clojure}, and the family of Clojure-like languages as
\emph{Clojure dialects}.

Clojure dialects are widely used for several reasons.
Performance is a key feature, with most dialects 
offering access to host-level performance

@@

When a static type checker is not available, which describes the situation
for most dynamic languages, other techniques are used for checking
type invariants. Runtime contracts and unit testing are a popular alternatives.
Clojure adopts these approachs, providing easy syntax for defining pre-
and post-conditions \footnote{Inspired by Eiffel} and a library for writing unit tests.
\begin{verbatim} but.. (why a static type system is still desirable\end{verbatim}

@@

@@

% Hello world
This section introduces Typed Clojure with example code. An attempt is
made to introduce some Clojure syntax and sematics to those unfamiliar or needing a refresher.
A basic knowledge of Lisp syntax is assumed.

@@

The \lstinline|clojure.core/ns|
\footnote{All vars from the \lstinline|clojure.core| namespace are referred implicitly in new namespaces (equivalent to \lstinline|(:require [clojure.core :refer :all])|).}
invocation declares the current namespace to be \lstinline|typed.test.hello-world|
and requires the namespace \lstinline|typed.core| to be loaded as a dependency, referring the var
\lstinline|typed.core/check-ns| to be referenced locally. Other than this
dependency, this is identical to the untyped \emph{Hello world}.
% footnote: check-ns used to initiate type checking usually at REPL

@@

(ann collatz [Number -> Number])
(defn collatz [n]
  (cond
    (= 1 n) 
      1
    (and (integer? n) 
         (even? n)) 
      (collatz (/ n 2))
    :else 
      (collatz (inc (* 3 n)))))
\end{lstlisting}
\footnote{\emph{Example adapted from Sam's dissertation TODO}}

In this example, we define a new var \lstinline|typed.test.collatz/collatz|. Typed Clojure requires all 
used vars to be annotated. Here \lstinline|typed.core/ann| annotates \lstinline|typed.test.collatz/collatz|
to be a function from \lstinline{java.lang.Number} to 
\lstinline{java.lang.Number}
\footnote{All Classes in the \lstinline|java.lang| package
are automatically imported in every Clojure namespace (the equivalent of Java's \lstinline|import java.lang.*;|).}.

>>>>

The Clojure programming language is a dynamically typed language running on the Java Virtual Machine. 
It emphasises functional programming with immutable data structures
and provides direct interoperability with languages on the Java Virtual Machine.
Clojure also is also a family of languages with similar semantics, referred to as Clojure dialects.
There are several Clojure dialects, notably the aforementioned Clojure for the Java Virtual Machine (JVM),
ClojureCLR which targets the Common Language Runtime, and ClojureScript which compiles to Javascript.
Where it is clear to do so, we refer to Clojure the programming language as \emph{Clojure}, 
and the family of Clojure-like languages as \emph{Clojure dialects}.

Clojure dialects are widely used for several reasons.
Performance is a key feature, for example the JVM implementation of Clojure
offers ways to access Java-like speed for certain operations.

@@

When a static type checker is not available, which describes the situation
for most dynamic languages, other techniques are used for checking
type invariants. Runtime contracts and unit testing are a popular alternatives.
Clojure adopts these approaches, providing easy syntax for defining pre-
and post-conditions \footnote{Inspired by Eiffel} and a library for writing unit tests.
\begin{verbatim} but.. (why a static type system is still desirable\end{verbatim}

@@

% Clojure community reactions
% GSOC
% Conj talk
% 
% Motivation: Conduit

@@

% Hello world
This section introduces Typed Clojure with example code. An attempt is
made to introduce some Clojure syntax and semantics to those unfamiliar or needing a refresher.
A basic knowledge of Lisp syntax is handy, but a brief tutorial is given
for newcomers.

\subsection{Preliminary: Lisp Syntax}

The core of understanding Lisp syntax can be summarised by these points

\begin{itemize}
  \item Invocations are always wrapped in a pair of balanced parenthesis 
  \item Parenthesis start to the left of the operator
\end{itemize}

For example, the Java expressions \emph{1 + 2 / 3} is written in Lisp pseudocode \lstinline|(/ (+ 1 2) 3)|
and \emph{numberCrunch(1, 2)} written \lstinline|(numberCrunch 1 2)|.

\subsection{Simple Examples}

@@

Here \lstinline|clojure.core/ns|
\footnote{All vars from the \lstinline|clojure.core| namespace are referred implicitly in new namespaces (equivalent to \lstinline|(:require [clojure.core :refer :all])|).}
declares the current namespace to be 
\lstinline|typed.test.hello-world|
and requires the namespace \lstinline|typed.core| to be loaded as a dependency,
referring \lstinline|typed.core/check-ns| to be referenced as \lstinline|check-ns|. Other than this
dependency, this is identical to the untyped \emph{Hello world}.
By running \lstinline|check-ns| we can verify that this namespace type checks.

@@

(ann collatz [Number -> Number])
(defn collatz [n]
  (cond
    (= 1 n) 
      1
    (and (integer? n) 
         (even? n)) 
      (collatz (/ n 2))
    :else 
      (collatz (inc (* 3 n)))))
\end{lstlisting}
\footnote{Example adapted from Tobin-Hochstadt \cite{Tob10}}

In this example, we define a new var \lstinline|typed.test.collatz/collatz|. Typed Clojure requires all 
used vars to be annotated. Here \lstinline|typed.core/ann| annotates \lstinline|typed.test.collatz/collatz|
to be a function from \lstinline{java.lang.Number} to 
\lstinline{java.lang.Number}\footnote{All Classes in the \lstinline|java.lang| package
are automatically imported in every Clojure namespace (the equivalent of Java's \lstinline|import java.lang.*;|).}.

### 2012-10-16 dissertation/lit-review.tex
Lots of stuff

Wright and Cartwright\cite{WC97} developed Soft Scheme, a practical
soft type system for Scheme. 
It extended earlier work by Cartwright and Fagan\cite{CF91}
and Fagan \cite{Fag92}, adding support for practical features such as
first-class continuations and variable-arity functions.
Soft Scheme does not require any extra type annotations.

% what challenges did soft typing face?

@@

Gradual typing combines static and runtime type checking so programmers
can the most appropriate one for the situtation.

@@

Clojure provides two dispatching mechanisms: multimethods, a mechanism to dispatch on the
results of an arbitrary function; and protocols, a mechanism for single dispatch.
Millstien and Chambers explore statically typing multiple dispatched multimethods in Dubious \cite{MS02}.
Dubious' multimethods are less powerful than Clojure's multimethods but more powerful than protocols,
and Clojure's dispatch functions can be extended at runtime, unlike Dubious' multimethods.

@@

Clojure provides mulimethods as a core language feature. This section discusses systems that statically
verify type safety for multimethods.

@@

>>>>

Wright and Cartwright\cite{WC97} developed Soft Scheme, a practical
soft type system for Scheme. 
It extended earlier work by Cartwright and Fagan\cite{CF91}
and Fagan \cite{Fag91}, adding support for practical features such as
first-class continuations and variable-arity functions.
Soft Scheme does not require any extra type annotations.

Type systems for Scheme have since moved away from soft typing
to to other approaches like gradual typing.
Alas, the best reference for commentary on this transition 
appears to be slides from a talk by Felleisen\cite{Fell09},
a leader in this area of research for over 20 years.
Felleisen says that while Soft Scheme discovered type problems, 
it suffers from incomprehensible
errors that required PhD-level expertise to decipher.

\begin{verbatim}
  Bidirectionality
\end{verbatim}

@@

Gradual typing combines static and runtime type checking so programmers
can the most appropriate one for the situation.

@@

Clojure provides two dispatching mechanisms: multimethods, a mechanism to dispatch on the
results of an arbitrary function; and protocols, a mechanism for single dispatch, a common
strategy that dispatches on the type of the first provided argument.
Millstien and Chambers explore statically typing multiple dispatched multimethods in Dubious \cite{MS02}.
Dubious' multimethods are less powerful than Clojure's multimethods but more powerful than protocols,
and Clojure's dispatch functions can be extended at runtime, unlike Dubious' multimethods.

@@

Clojure provides multimethods as a core language feature. This section discusses systems that statically
verify type safety for multimethods.

@@

\section{Higher Kinded Programming}

\begin{verbatim}
Many advanced type systems provide support for ..
Forward reference to algo.monad experiment.
\end{verbatim}

### 2012-10-16 dissertation/paper.tex
Lots of stuff

\usepackage{mmm}
\usepackage{mathpartir}
\usepackage{clj-grammar}
\usepackage[style=numeric]{biblatex}
\addbibresource{bibliography.bib}

@@

\begin{abstract}
  % FIXME long tradition for attempts
  There is a long tradition to create static type systems for dynamically-typed
  languages. Static type checking promises to eliminate many common user errors
  that are otherwise unnoticed or are caught when the program is run.

  Recent attempts have seen several promising and innovative
  type systems built for dynamically typed languages. Typed Racket is able to
  statically type check idioms ..
  We take the lessons learnt from these
  projects and apply them to the dynamically-typed language Clojure in the form of Typed Clojure.
\end{abstract}

%\tableofcontents

@@

>>>>

\usepackage{mmm}
\usepackage{mathpartir}
\usepackage{clj-grammar}
\usepackage[style=alphabetic]{biblatex}
\addbibresource{bibliography.bib}

@@

\begin{abstract}

Traditionally there has been a clear distinction between 
programming languages that perform type checking at compile-time, and those that do not.
In particular, the early wide-spread programming languages ALGOL and LISP
could be divided this way: ALGOL performs compile-time (or static) type checking,
and LISP does not.
Static type checking eliminates many common user errors that are otherwise unnoticed or are caught 
when the program is run.
On the other hand, a language without static type checking can often be designed 
to be more flexible and forgiving.

A recent trend is to combine the advantages of both kinds of languages by adding optional static 
type systems to languages without static type checking. 
A major challenge in these combinations is continuing to support the idiomatic style of the original 
language, and some recent attempts appear to have finally produced type systems that have overcome this obstacle.  
Most notably Typed Racket, a typed sister language of the Lisp dialect Racket, is able to statically type 
check most idiomatic Racket code.

Clojure is a recent and increasingly popular programming language that is designed to support 
functional programming and immutability by default. Implementations of Clojure are hosted 
on popular platforms and provide direct interoperability with the host platform.
There are several implementations of Clojure, each targeting different runtimes including
the Java Virtual Machine, the Common Language Runtime, and Javascript virtual machines.

This paper describes my work on designing an optional static type system for Clojure based on
the lessons learnt from several projects, primarily Typed Racket.
A prototype type system is developed for Clojure running on the Java Virtual Machine.
This prototype is sufficient to allow experiments with existing Clojure code that is sufficiently complicated
that type checking increases confidence that the code is correct.
For example, a significant portion of algo.monads, a Clojure Contrib library for monadic programming, 
is able to be type checked. 
On the positive side, most monad, monad transformer, and monadic function definitions can be type checked,
usually by adding type annotations in natural places like function definitions.
Negatively, simple monad comprehensions can be type checked, but require awkward type annotations.

There is significant future work to fully type check all Clojure features and idioms.
For example, multi-method definitions and functions accepting an even number of variable arguments
are troublesome. 
Also, there are desirable features from the Typed Racket project that are missing, such
as automatic runtime contract generation and a sophisticated blame system, 
both which significantly improve error messages when mixing typed and untyped code.
I conclude that it appears to be both practical and useful to design and implement an optional static type system for the
Clojure programming language.

\end{abstract}

\tableofcontents

@@

\chapter{Conclusion}

\begin{verbatim}
  Programming is an error-prone task, and verification tools are important.
  Dynamic languages rely on several techniques for verification, but static
  type checking is not often supported.
\end{verbatim}
